\documentclass[11pt, oneside]{amsart}
\pdfoutput=1

\usepackage{amsmath,amssymb,amsmath}

\usepackage{color}
\usepackage[table]{xcolor}
\usepackage{dcolumn}
\usepackage{caption}
\usepackage{graphicx}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[font=small]{caption}
\usepackage{varwidth}

\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\microtypecontext{spacing=nonfrench}
\usepackage{xspace}

\usepackage{rotating}
\usepackage{caption,subcaption}
\usepackage{float}
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}
\usepackage{longtable}
\usepackage{verbatim}
\usepackage{array,booktabs,multicol,multirow}

%better printing of numbers
\usepackage[english]{babel}
\usepackage{textcomp}
\usepackage{csquotes}

% The following three lines are used for displaying footnote in tables.
\usepackage{footnote}
\makesavenoteenv{tabular}
\makesavenoteenv{table}


\usepackage{enumitem}
\setlist{leftmargin=7mm}

%\setcounter{secnumdepth}{3}
%\setcounter{tocdepth}{3}

\usepackage[bookmarks, bookmarksopen, bookmarksnumbered]{hyperref}
\usepackage[all]{hypcap}
\urlstyle{rm}

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\definecolor{darkgreen}{rgb}{0,0.6,0}
\definecolor{cyan}{rgb}{0.2,0.7,0.7}
\definecolor{blueish}{rgb}{0.2,0.2,0.8}
\definecolor{darkblue}{rgb}{0.1,0.1,0.9}
\definecolor{lightgray}{gray}{0.9}

\newcommand{\todo}[1]{{\color{blue}$\blacksquare$~\textsf{[TODO: #1]}}}
\newcommand{\note}[1]{ {\textcolor{blueish}    { ***Note:      #1 }}}
\newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
\newcommand{\gabnote}[1]{ {\textcolor{cyan}    { ***Gabrielle:     #1 }}}
\newcommand{\nchnote}[1]{  {\textcolor{orange}      { ***Neil: #1 }}}
\newcommand{\kylenote}[1]{  {\textcolor{cyan}      { ***Kyle: #1 }}}

% Don't use tt font for urls
\urlstyle{rm}

\usepackage[letterpaper, margin=1in]{geometry}
% You can use a baselinestretch of down to 0.9
\renewcommand{\baselinestretch}{0.96}

\sloppypar

\begin{document}

\title[]{Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)}

\author{Daniel~S.~Katz$^{(1)}$,
Kyle~E.~Niemeyer$^{(2)}$,
Sandra~Gesing$^{(3)}$,
Lorraine~Hwang$^{(4)}$,
Simon~Hettrick$^{(5)}$,
Alice~Allen$^{(6)}$,
Jean~Salac$^{(7)}$,
Neil~Chue~Hong$^{(8)}$,
Santiago~Nunez-Corrales$^{(9)}$
Anshu~Dubey$^{(10)}$,
R.~Stuart~Geiger$^(11)$,
Jonah~Miller$^{12)}$,
Emily~Chen$^{123}$,
Wolfgang~Bangerth$^{14)}$,
\\Author order to be determined based on contributions
}

%
\thanks{{}$^{(1)}$ \hspace{-1ex}National Center for Supercomputing Applications (NCSA) \&
Department of Computer Science  \&
Department of Electrical and Computer Engineering  \&
School of Information Sciences (iSchool),
University of Illinois, Urbana--Champaign, IL, USA; d.katz@ieee.org; ORCID: 0000-0001-5934-7525}
%
\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering,
Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: 0000-0003-4425-7097}
%
\thanks{{}$^{(3)}$ Center for Research Computing \& Department of Computer Science and Engineering,
University of Notre Dame, IN, USA; sandra.gesing@nd.edu; ORCID: 0000-0002-6051-0673}
%
\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu; ORCID: ?}
%
\thanks{{}$^{(5)}$ The Software Sustainability Institute, Southampton, UK; s.hettrick@software.ac.uk; ORCID: 0000-0002-6809-5195}
%
\thanks{{}$^{(6)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
%
\thanks{{}$^{(7)}$ \note{info needed}}
%
\thanks{{}$^{(8)}$ \note{info needed}}
%
\thanks{{}$^{(9)}$ \note{info needed}}
%
\thanks{{}$^{(10)}$ \note{info needed}}
%
\thanks{{}$^{(11)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
%
\thanks{{}$^{(12)}$ \note{info needed}}
%
\thanks{{}$^{(13)}$ \note{info needed}}
%
\thanks{{}$^{(14)}$ Department of Mathematics, Colorado State
  University, Fort Collins, CO 80523, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}


\begin{abstract}
This report records and discusses the Fourth Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE4). The report
includes a description of the keynote presentation of the workshop,
which % TODO
% It also summarizes a set of lightning
% talks in which speakers highlighted to-the-point lessons and challenges
% pertaining to sustaining scientific software.
% The final and main contribution of the report is a summary of the
% discussions, future steps, and future organization for a set of self-organized
% working groups on topics including developing pathways to funding scientific
% software; constructing useful common metrics for crediting software
% stakeholders; identifying principles for sustainable software engineering
% design; reaching out to research software organizations around the world; and
% building communities for software sustainability. For each group, we include a
% point of contact and a landing page that can be used by those who want to join
% that group's future activities. The main challenge left by the workshop is to
% see if the groups will execute these activities that they have scheduled, and
% how the WSSSPE community can encourage this to happen.

\end{abstract}



\maketitle
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Fourth Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
held over 2 1/2 days on 12--14 September 2016 in Manchester, England, UK.
This location and date was selected so that WSSSPE4 immediately preceded the
First Research Software Engineers (RSE) Conference.
Previous events in the WSSSPE series include
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
held in conjunction with SC13;
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
focused workshop organized jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
held in conjunction with SC14;
WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
focused workshop organized again jointly with
SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}};
and WSSSPE3\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}}~\cite{WSSSPE3},
held in Boulder, Colorado, USA.
Note that the first WSSSPE workshop was named
``Working towards
Sustainable Software for Science: Practice and Experiences,'' which remains the meaning
of the WSSSPE group, but the workshops after that were named
``Workshop on Sustainable
Software for Science: Practice and Experiences.'' Together these reflect
that WSSSPE is both a community and a set of workshops.

%\todo{Introduction}

The WSSSPE4 workshop included multiple mechanisms for participation and
encouraged team building around solutions. WSSSPE4 strongly encouraged participation
of early-career scientists, postdoctoral researchers, graduate students,
early-stage researchers, and those from underrepresented groups,
with funds provided to the conference organizers by the National Science
Foundation (NSF), the Gordon and Betty Moore Foundation, the Alfred P.~Sloan Foundation, and the Software
Sustainability Institute (SSI) to support the travel of potential participants
who would not otherwise be able to attend the workshop. These
funds allowed 29 additional people to attend and participate. A subset of the
organizing committee reviewed 44 applications for travel support and
competitively selected the awardees, including 11 students and five early-career
researchers. In addition, the travel award subcommittee tried to increase the
diversity of applicants by forwarding the notice of travel support to
organizations including PyLadies\footnote{\url{http://www.pyladies.com}},
Django Girls\footnote{\url{https://djangogirls.org}},
Women Who Code\footnote{\url{https://www.womenwhocode.com}}, and
Women in HPC\footnote{\url{http://www.womeninhpc.org}}.
\todo{did we reach out to any others?}

WSSSPE4 also included two professional event organizers/facilitators from
KnowInnovation who helped the organizing committee plan the workshop agenda,
and during the workshop, they actively engaged participants with various
tools, activities, and reminders.

This workshop also introduced a Code of Conduct (CoC)
\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
for both WSSSPE4 and interactions in the greater WSSSPE community (e.g., on
Twitter, in email lists, in the Slack team). The WSSSPE4 CoC is based on the
FORCE11 conference CoC~\cite{FORCE11:CoC}, in turn based on the Code4Lib
CoC~\cite{Code4Lib:CoC}.
The main guidelines of the CoC are:
\begin{quote}
    WSSSPE events are community events intended for networking and collaboration
    as well as learning. We value the participation of every member of the
    community and want all attendees to have an enjoyable and fulfilling
    experience. Accordingly, all attendees are expected to show respect and
    courtesy to other attendees throughout the event and in interactions online
    associated with the event.

    The WSSSPE event organizers are dedicated to providing a harassment-free
    experience for everyone, regardless of gender, gender identity and
    expression, age, sexual orientation, disability, physical appearance,
    body size, race, ethnicity, religion (or lack thereof), technology choices,
    or other group status.

    To make clear what is expected, everyone taking part in WSSSPE events and
    discussions---speakers, helpers, organizers, and participants---is required
    to conform to the following Code of Conduct.

    \begin{itemize}
    \item Communicate appropriately for a professional audience including
    people of many different backgrounds. Sexual language and imagery are not
    appropriate for any event.

    \item Be kind to others. Do not insult or put down other attendees. Be
    mindful of jargon, which can sometimes exclude others from engaging in the
    discussion.

    \item Behave professionally. Remember that harassment and sexist, racist,
    ageist, or exclusionary behavior are not appropriate.
    \end{itemize}
\end{quote}

The CoC was introduced at the beginning of WSSSPE4, with the CoC subcommittee
and a general email address introduced for reporting concerns or incidents, or
asking questions.  There was one concern mentioned to the CoC subcommittee
after the first half day,
regarding how part of the workshop was being run, and we changed the workshop
to address this.

This report is based on the events at the workshop and the submitted materials.
There events included discussion of the WSSSPE mission and vision, a keynote,
a set of presentations, a panel, and a number of working groups.  About a day of the
workshop was spent with participants in the working groups, which occurred in parallel
with each other.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Calls for participation} \label{sec:preworkshop}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

WSSSPE4 was based on the work done in WSSSPE1, WSSSPE2, and WSSSPE3, but aimed
at producing working groups that better continued working after the workshop ended.
In addition, based on feedback after WSSSPE3, it became clear that WSSSPE attendees
had two different motivations in participating.  One motivation was to make a better future
for research software, and the other was to immediately do better research software development.
This led to the idea of WSSSPE4 being partially divided into two tracks:

\begin{quote}
    \textbf{Track 1 -- Building a sustainable future for open-use research
    software} has the goals of defining a vision of the future of open-use
    research software, and in the workshop, initiating the activities that are
    needed to get there. The idea of this track is to first think about where
    we want to be 5 to 10 years from now, without being too concerned with
    where we are today, and then to determine how we can move to this future.

    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}
    has the goal of improving the quality of today's research software and the
    experiences of its developers by sharing practices and experiences.
    This track is focused on the current state of scientific software and what
    we can do to improve it in the short term, starting with where we are today.
\end{quote}

The call for participation requested
idea papers, which present implementable proposals related to Track 1;
position papers, which are longer, not previously published papers related to
Track 2 specifically discussing what we can do to improve sustainable scientific
software in the short term, starting with where we are today;
experience papers, longer papers related to Track 2 that discuss current
practices and experiences and how they have been used to improve the quality of
today's research software and/or the experiences of its developers;
demos, brief papers describing a tool or
process that would be demonstrated, relevant to Track 2 that improves the quality of today's research
software and\slash or the experiences of its developers; and
extended abstracts describing lightning talks,
where each author could make a brief statement about work that either had been
done or was needed.

The following list of contribution topics was suggested, though this was not intended to limit potential submissions.

%
%\begin{quote}
\begin{itemize}
\renewcommand{\labelenumi}{\textbf{\theenumi}.}
\setlength{\rightmargin}{1em}

\item Development and Community
\begin{itemize}
    \item Best practices for developing sustainable software
    \item Models for funding specialist expertise in software collaborations
    \item Software tools that aid sustainability
    \item Academia/industry interaction
    \item Refactoring/improving legacy scientific software
    \item Engineering design for sustainable software
    \item Metrics for the success of scientific software
    \item Adaptation of mainstream software practices for scientific software
\end{itemize}

\item Professionalization
\begin{itemize}
    \item Career paths
    \item RSE as a brand
    \item RSE outside of the UK or Europe
    \item Increase incentives in publishing, funding and promotion for better software
\end{itemize}

\item Training
\begin{itemize}
    \item Training for developing sustainable software
    \item Curriculum for software sustainability
\end{itemize}

\item Credit
\begin{itemize}
    \item Making the existing credit and citation ecosystem work better for software
    \item Future credit and citation ecosystem
    \item Software contributions as a part of tenure review
    \item Case studies of receiving credit for software contributions
    \item Awards and recognition that encourage sustainable software
\end{itemize}

\item Software publishing
\begin{itemize}
    \item Journals and alternative venues for publishing software
    \item Review processes for published software
\end{itemize}

\item Software discoverability/reuse
\begin{itemize}
    \item Proposals and case studies
\end{itemize}

\item Reproducibility and testing
\begin{itemize}
    \item Reproducibility in conferences and journals
    \item Best practices for code testing and code review
\end{itemize}

\end{itemize}

Submissions to WSSSPE4 comprised
19 lightning talks,
4 idea papers,
3 position paper,
5 experience papers,
and
3 demos.
Most of the submissions were accepted, since the goal of WSSSPE is always to
take in as many relevant inputs as possible, and to encourage their authors to
participate in sharing and implementing their ideas.
Specifically,
19 lightning talks,
4 idea papers,
3 position papers,
2 experience papers,
and
3 demos
were accepted. (Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
were accepted as lightning talks.)
The papers are discussed in Section~\ref{sec:papers},
and the lightning talks are discussed in Section~\ref{sec:lightning}.
The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynote}\label{sec:keynote}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\note{need a person to write this - Dan has asked Patricia if she will, if not, perhaps Dan will do it}

The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Mission and vision}\label{sec:mission}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\note{by Lorraine and Dan}

Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
The organizers developed a strawhorse, which was presented to the participants early in the meeting.
The presentation included guidelines as well as examples from other similar communities.
The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; Its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.

The participants were given time after the presentation to write down their comments.
Seven people volunteered to work on revising the mission and vision statements, incorporating these comments and presenting back to the community.
Based on this feedback, the committee redrafted the mission and vision.
The committee added focus areas to clarify the organizations breadth and to keep the mission and vision simple and long-lasting.
Comments after the close of the meeting were incorporated into the statements.

A final draft was put on GitHub (\url{https://github.com/WSSSPE/mission-vision}) and advertised
via the WSSSPE mailing list, Facebook group, and Twitter.
After 2 weeks without suggested changes, the final statements, as shown below, were added
to the WSSSPE web page (\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}).

{\bf Mission}
WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.

{\bf Vision}
We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.

{\bf Focus areas}
WSSSPE promotes sustainable research software by positively impacting:
\begin{itemize}
\item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
\item {\bf Careers}. Developing and supporting career paths in research software development and engineering
\item {\bf Learning}. Engaging in activities to promote peer learning and interaction
\item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
\end{itemize}

{\bf Definitions}

Sustainable software has the capacity to endure such that it will continue to be available in the future, on new platforms, meeting new needs.

The research software lifecycle includes:
\begin{itemize}
\item acquiring and assembling resources (including funding and people) into teams and communities
\item developing software
\item using software
\item recognizing contributions to and of software
\item maintaining software
\end{itemize}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Papers and demos} \label{sec:papers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\note{Sandra Gesing will write this}

WSSSPE included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
two experience papers, and three demos) that addressed a wide range of topics around
sustainability for software in science. They covered three main areas:
\begin{enumerate}
\item The specifics of the academic environment for software development and for so-called
research programmers or research software engineers;
\item The characteristics and needs of research software in general such as the lifecycle of
software in science, preservation, curation, sharing, and reproducibility challenges; and
\item Lessons learned from or visions for use cases of scientific software and communities
working with and\slash or on a software package.
\end{enumerate}

\todo{group by type - idea, position, experience, demo}
\begin{itemize}
\item \textbf{Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
Improving Incentives for Quality Software}}~\cite{Heroux:2016ws}.
The author presents that improving quality of scientific software is a measure
to address the increased complexity in both modeling and software. To support
quality, it is essential to provide Computational Science and Engineering teams
with publication venues, funding, and  professional recognition. Broad
improvements in sustainability and productivity can be only achieved when
publishers, funding agencies and employers raise their expectations for
software quality. Additionally, software community leaders are in the unique
position to positively impact software quality by working on establishing
incentives that will spur creative and novel approaches to improve developer
productivity  and software sustainability.

\item \textbf{Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
Society for Research Software}}~\cite{GAllen:2016ws}.
This paper advocates to create a professional society or extend the scope of an
existing organization in research software. The author presents the motivation,
a variety of potential roles of such a society and diverse potential funding
mechanisms to be able to establish a sustainable organization. She gives an
overview on similar activities as well as existing organizations and aims at
stimulating discussion among potential members and stakeholders.

\item \textbf{Olivier Philippe, Simon Hettrick, and Neil Chue Hong:
\emph{Experience Paper: Preliminary analysis of a survey of UK Research
Software Engineers}}~\cite{Philippe:2016ws}.
This paper presents results from a survey conducted on the new role of research
software engineers in academia. The survey provides demographic information
about the education, field, gender, job satisfaction and career plans of the people
in this community. The outcome is that members of the community  were found to be
highly educated, derive mainly from the hard sciences, and are predominantly male.
Respondents reported satisfaction in their jobs, but indicated that career progression
is both difficult and opaque. The authors urge a continued discussion about the experience
of research software engineers and recommend further investigation into this
important community.

\item \textbf{Christopher Gwilliams: \emph{Demo: Using industrial engagement to
create and develop research ties within academia}}~\cite{Gwilliams:2016ws}.
This demo focuses on the benefits of including a  new  software engineering
degree that is influenced by, and works directly with,
industry.  Cardiff  University  introduced  a  new  industry-focused  degree,
Bachelor of Science in Applied Software Engineering (ASE) in September
2015. The engagement with industry allows students to learn through live
projects  with  real-world  applications and forge links for career opportunities.
Also, industry benefits from this course that provides an entry point for industry
to engage with the university and to develop a network of contacts between industry
and academia, which is also a benefit for the university and allows to identify
potential opportunities for new research projects and collaborators.

\item \textbf{Anshu Dubey and Katherine Riley: \emph{Experience Paper: Software
Engineering and Community Codes Track in ATPESC}}~\cite{Dubey1:2016ws}.
The Argonne Training Program in Extreme Scale
Computing (ATPESC) was started by Argonne National Laboratory with the objective
of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
One of the unique aspects of the program was inclusion of a
software engineering and community codes track. The inclusion
was motivated by the observation that the projects with a good
software process were better able to meet their scientific goals.
The paper presents the experience in running the software
track, the motivation, reception, and evolution of the track over the years.
The authors recommend the inclusion of similar tracks in other HPC-oriented training programs.

\item \textbf{Neil Chue Hong: \emph{Position Paper: Why do we need to compare
research software, and how should we do it?}}~\cite{ChueHong:2016ws}.
This position paper sets out the reasons for why different stakeholders,
from users to developers to funders, might wish to undertake the difficult task
to compare research software. The author describes a proposed framework for doing
so (based around measures of accessibility, usability, maintainability and
portability) which takes into account the possibility of variation between
different communities about how they prioritize different aspects of research software.

\item \textbf{Anshu Dubey and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
of Software for Scientific Simulation}}~\cite{Dubey2:2016ws}.
The authors discuss the specifics of the lifecycle of scientific software.
While the software lifecycle is a well-researched topic, one class of projects,
research software, has received relatively
little attention from lifecycle researchers. This
paper formalizes a lifecycle model
for end-to-end scientific application software, featuring loose
coupling between submodels for development of infrastructure
and scientific capability. The authors also invite input from stakeholders
to converge on a model that captures the complexity of this
development processes and provides needed lifecycle guidance to
the scientific software community.

\item \textbf{Francisco Queiroz and Rejane Spitz: \emph{Position Paper:
Collaborative Gamification Design for Scientific Software}}~\cite{Queiroz:2016ws}.
The paper focuses on usability design and presents that gamification, a design
trend for improving scientific software usability, for scientific software should
be facilitated by an open, collaborative design process supported by
conversational media. This approach is compatible with
qualities often attributed to computational science community
regarding openness and collaboration between members of varied
professional backgrounds. They exemplify the use of conversational media for
collaborative design and expect the synergy between collaborators to result in
better usability, greater user acceptance, and adequacy to requirements,
obtaining optimal design solutions in a sustainable way.

\item \textbf{Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous, and
David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
the Way Digital Libraries will Operate}}~\cite{Childers:2016ws}.
The paper presents the implications of ``active curation systems'' that provide
executable access to
artifacts and experiments behind published results and enables
their reuse. These systems allow changing and repeating experiments to understand
how an innovation behaves in conditions beyond the ones described in a paper and
address reproducibility challenges. Active curation systems
also  enable accountability and accelerate research progress  by
giving access to complete experimental details. As these systems
take  hold,  it  is  important  to  understand  their  capabilities  and
how digital libraries should be integrated with them. The authors describe a
study underway to explore how best
to do this integration.

\item \textbf{Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse: \emph{Demo: Composing,
Reproducing, and Sharing Simulations}}~\cite{Ganguly:2016ws}.
The authors demonstrate their approach of sharing,
reproducing, and composing simulations toward accelerating
research productivity while also improving accountability and
credibility. Specifically, they have developed a case study in
which they compose and share access control simulations in the
form of shareable data store units for cloud systems. They share, compose
and repeat simulations  through a collaborative repository
(OCCAM) and suggest a general simulation framework (SST) that can
accelerate the efforts for the whole community.

\item \textbf{Sameer Shende and Allen Malony: \emph{Demo: Using TAU for
Performance Evaluation of Scientific Software}}~\cite{Shende:2016ws}.
This demo presents the  TAU Performance System for performance evaluation of Scientific
Software written in C++, C, and Fortran. The system addresses performance
technology problems  at three levels: instrumentation,
measurement, and analysis. The TAU framework supports the
configuration and integration of these layers. The authors advocate that effective
exploration of performance will necessarily
require prudent selection from the range of alternative methods TAU provides
to  assemble meaningful performance experiments.

\item \textbf{Janos Sallai, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
Atom-Typing for Molecular Simulation}}~\cite{Sallai:2016ws}.
The authors present an essential and error-prone task in the area of Molecular
Dynamics simulations: the efficient selection of correct parameter values for
specific molecules, especially forcefields. They present a framework that
aims to solve this data management issue, proposing a common format for
forcefields that is self-documenting with machine readable,
declarative usage rules. They investigate processes and tools
that are commonly used today in software development (e.g.,
unit testing, verification and validation, continuous integration,
and version control) and reason that they are, with proper infrastructure
support, applicable to forcefield development, as well. The
paper describes how such an infrastructure can tackle managing and evolving
forcefields by the molecular dynamics community, and
proposes a way to encourage and incentivize involvement
by the stakeholders.


\end{itemize}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{
\href{http://wssspe.researchcomputing.org.uk/wssspe4/agenda/}{Slides.}}
\end{comment}

There were 19 lightning talks presented at WSSSPE4, as follows (in order of
presentation).
%
A \textsuperscript{\textasteriskcentered} indicates which author presented the
lightning talk, for those with multiple authors.
%
\begin{itemize}[itemsep=1ex]
    \item \textbf{Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
    programme; supporting the social side of research software}}~\cite{Sufi:2016ws}.
    Sufi, Community Lead at the Software Sustainability Institute (SSI), discussed
    the SSI Fellowship program and the fellows, including their
    demographic breakdown and domains of practice. This program came about
    based on the institute's desire to integrate with the research community---thus
    inspiring change from within---rather than preach to it. Five SSI fellows
    attended WSSSPE4.

    \item \textbf{Sandra Gesing, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
    Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
    \emph{The Science Gateways Community Institute - Supporting Communities to
    Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
    In this talk, Gesing introduced the concept of science gateways: end-to-end software
    solutions for a community that hide underlying complexity. She also posed the need
    for gateways, which include the increased complexity of research and associated
    hardware\slash software, a greater need for reproducibility and openness, and the
    opportunity to integrate research with teaching for better workforce development.
    In fact, gateway users surpassed login-based users in XSEDE systems in 2013,
    and 1.4 times as many users accessed XSEDE via gateways compared with
    login in March 2016. The Science Gateways Community Institute offers support
    related to gateways in five areas: Incubator, Extended Developer Support, Scientific
    Software Collaborative, Community Engagement and Exchange, and Workforce
    Development.

    \item \textbf{Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
    and Scott Chamberlain: \emph{A model for
    peer review and onboarding research software}}~\cite{Ram:2016ws}.
    This talk described the rOpenSci community's review process for research
    software packages. Submission of a package starts with a pre-submission inquiry
    for fit with the community criteria, then peer reviewers evaluate based on
    usability, quality, and style. Upon acceptance, packages are given an rOpenSci
    README badge and added to the system. Important components include a code of
    conduct, open and non-adversarial review, and that no submission is
    rejected---just out of scope. Remaining challenges include scaling, author
    incentives, and automation of the process.

    \item \textbf{Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
    \emph{A vision of computing in 10+ years}}~\cite{Loffler:2016ws}.
    This talk began by describing the challenges facing science, and particularly
    computational science, in recent years: the amount and complexity of knowledge
    has increased dramatically, making it difficult for individuals or even teams to
    manage modern, complex problems.
    Brandt submitted an argument for decoupling science simulation and
    the solution of computational problems away from numerical methods and low-level
    optimization. This may be possible with domain-specific languages, or the use
    of modern features in existing languages. This vision also includes reuse of
    basic computing infrastructure, including common data structures, parameter files,
    and high-performance I\slash O. This computing future necessarily requires
    better recognition and reward systems for research software developers (i.e.,
    research software engineers), and ways to make increasingly complex scientific
    papers more understandable.


    \item \textbf{Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
    Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
    \emph{Software Citation: Process, Principles, and Implementation}}~\cite{Katz:2016ws}.
    This talk described the timeline of activities of the FORCE11 Software Citation
    Working Group and the principles of software citation established by the group.
    Niemeyer, one of the working group cochairs, also discussed the paper published
    describing the principles~\cite{Smith:2016sc}, and the remaining steps of the
    working group before sunsetting it and spinning up a new Software Citation
    Implementation Working Group.

    \item \textbf{Ray Idaszak\textsuperscript{\textasteriskcentered},
    David G.~Tarboton, Hong Yi, Michael Stealey,
    Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
    Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
    Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
    Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
    Best Practices and Culture Change for Developing Sustainable Community
    Software}}~\cite{Idaszak:2016ws}.
    HydroShare is a five-year-old online, collaborative system that supports
    open sharing of hydrologic data, analytical tools, and computer models. The
    project involves 10 collaborating institutions that produce tested and reviewed
    community code every two to three weeks, following software engineering best
    practices. Undergraduate and graduate students involved in the project not
    only adopt the best practices exhibited by HydroShare, but see the open,
    collaborative approach to research software development as the norm.
    The talk made recommendations to other community efforts, including the need to
    identify committed, hands-on PIs; evidence-based software engineering for science;
    continuous integration; the importance of professional software development and
    operations practices, even for domain scientists; and to identify technology
    decisions that overlap with career paths.

    \item \textbf{Lorraine Hwang\textsuperscript{\textasteriskcentered},
    Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
    \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
    Community}}~\cite{Hwang:2016ws}.
    In this talk, Hwang first described the ASPECT (Advance Solver for Problems in
    Earth ConvecTion) open-source, parallel, extensible finite element code
    used to simulate thermal convection. ASPECT began development in 2011 as a
    community-driven project under the Computational Infrastructure for Geodynamics
    (CIG). Community-building activities around ASPECT focused on the traditional
    workshops, tutorials, and webinars, but also on slightly nontraditional,
    multi-day hackathons---which turned out to be one of the most effective
    forms of fostering a community. These start with leaders establishing project
    conventions, demonstrating workflows and relevant technical skills, and
    identifying starter projects. Daily activities involve work plans and
    science\slash coding briefs, with mini-tutorials given as needed. One important
    requirement is that each participant must submit at least one commit. Hwang
    also summarized lessons learned, including that hackathons should last 7--10
    days, include participants with multidisciplinary expertise and at both novice
    and expert levels, have at least one dedicated code reviewer per 10 participants,
    hold daily group updates, organize activities to promote group interactions,
    and enforce free time.

    \item \textbf{Carole Goble: \emph{A Simple Profiling Framework for Software
    User-Producer Reciprocity Review}}~\cite{Goble:2016ws}.
    Goble first described some existing frameworks for measuring the maturity of
    software, including the Software Sustainability Maturity Model, OSS Watch
    Openness Rating, NASA Reuse Readiness Rating, Capability Maturity Model, and
    Qualification and Selection of Open Source software. Then, she proposed a new,
    simpler framework for project reciprocity, drawing on the intentions of users
    and producers of the software, from the viewpoint of the producer.
    Next steps include further definition of the model's characteristics based on
    both theoretical and empirical investigation.

    \item \textbf{Stephan Druskat: \emph{A proposal for the measurement and
    documentation of research software sustainability in interactive metadata
    repositories}}~\cite{Druskat:2016ws}.
    In this talk, Druskat first identified two technical barriers to software
    sustainability: identification of good software, and software discoverability.
    Some of his perspectives were based on experiences in the humanities, a lack
    of funding for computational infrastructure results in little experience with
    sustainability (although this may differ in other domains). Druskat proposed
    metadata repositories as a possible solution, as they can measure and document
    technical sustainability, and document resulting metrics and more general
    features of software. Technical sustainability goals include ensuring software
    existence, preserving potential operation of software, and creating\slash
    retaining possibilities for further development and maintenance of software.
    The next steps of such a system include compiling measurable criteria for
    the sustainability goals (based on qualitative concepts of, e.g.,
    ``sustainability,'' ``maintainability,'' and ``usability''), and
    crowd-sourcing a weighting of these criteria.

    \item \textbf{Mistral Contrastin, Matthew Danish,
    Dominic Orchard\textsuperscript{\textasteriskcentered}, and
    Andrew Rice: \emph{Supporting software sustainability with
    lightweight specifications}}~\cite{Contrastin:2016ws}.
    This talk introduced the tool CamFort for verifying Fortran code based on
    lightweight specifications, which aid in software maintainability and reuse
    through high-level indications of programmer intent.
    Currently, CamFort supports specification (and
    verification) of units and stencil operations. Plans for future specifications
    include automatic test generation based on user-supplied properties, dependency
    specifications, and software contracts to check (e.g.) expected value ranges
    and program behavior.

    \item \textbf{Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
    Andrei Chi\c{s}, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
    Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
    Robert Haines, Robert Hirschfeld, James Howison,
    Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
    Ralf L\"{a}mmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
    and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
    Responsibility in the Engineering of Academic Software}}~\cite{AAllen:2016ws}.
    This talk described the activities of the weeklong Dagstuhl Perspectives Workshop
    manifesto on ``Engineering Academic Software,'' which produced a manifesto
    describing a pledge affirming the scholarly value of research software and
    the personal responsibility to act in accordance with that value. The draft
    (personal) manifesto includes 20 pledges, organized into three areas:
    recognition of academic software, academic software development processes,
    and the intellectual content of academic software. The talk concluded with a
    survey for participants to give feedback on the pledges and their importance.

    \item \textbf{Neil Chue Hong\textsuperscript{\textasteriskcentered}:
    \textit{Making it easier to understand research software impact}}~\cite{ChueHong:2016wsb}.
    Chue Hong proposed that the WSSSPE community collaboratively develop a framework
    to better understand (and measure the impact of) research software. This could
    be based on a workflow involving mining the literature for mentions of software,
    identifying linkages between software, and visualizing dependencies between
    software or between software and other research artifacts. Many tools already
    exist in this ecosystem, so Chue Hong proposed that a WSSSPE working group
    collect these tools, summarize use cases, identify missing pieces, and find
    low-hanging fruit to integrate tools or build new ones.

    \item \textbf{Willem Robert van Hage\textsuperscript{\textasteriskcentered},
    Jason Maassen and Rob van Nieuwpoort: \textit{Software Impact Measurement at the
    Netherlands eScience Center}}~\cite{vanHage:2016ws}.
    This talk described eStep (eScience Technology Platform) at the Netherlands
    eScience Center, which includes three components: a software catalogue
    of eScience software, interfaces, libraries,
    tools workflows, applications, etc.; a knowledge base of guides, reports, and
    recommendations for scientific software development; and eScience research
    documents including such as scientific publications, book chapters, and demos.
    The eStep website currently presents automatically-generated information about
    software based on metadata, with a plan for displaying measures of impact and
    quality.

    \item \textbf{Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
    Ray Idaszak, and Daniel S.\ Katz: \textit{Report on Software Metrics for
    Research Software}}~\cite{GAllen:2016wsb}.
    In this talk, Chen first summarized some of the outcomes of the WSSSPE3
    Metrics Breakout Group, who discussed that metrics are important for scientific
    impact, discovery, reducing duplication of efforts, and evaluating software.
    The group also summarized barriers to widespread use of metrics: no common
    set of metrics collected, a need for usefulness to be obvious for projects,
    and no common standard for collecting\slash presenting metrics. To find out
    more about practices related to metrics, the authors ran a survey of National
    Science Foundation Software Infrastructure for Sustained Innovation (SI2)
    program investigators, with 19 participants. Important results include the
    observation that most use citations and acknowledgements of software in
    funding proposals, while download numbers are both challenging to collect
    and not useful. Overall, participants generally agreed that metrics are
    important, but said useful metrics are difficult or tedious to collect.

    \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
    and Gabrielle Allen: \textit{Bringing Techniques from Software Engineering
    into Scientific Software}}~\cite{Seidel:2016ws}.
    This talk, given by a senior graduate student in Computer Science\slash
    programming languages research, described several areas of software engineering
    concepts that benefit the development of research software. In particular,
    Seidel focused on software correctness, starting from high-level
    specifications. Correctness can be demonstrated by verifying or proving
    that the software satisfies the specification, testing against a range of
    inputs, or synthesis of code from the specification. However, all of these
    present technical challenges, in addition to the social challenges of
    encouraging collaboration between software engineering and research software
    communities.

    \item \textbf{Iain Emsley\textsuperscript{\textasteriskcentered} and David
    De Roure: \textit{Sustaining the social: Connecting the lives of Drupal
    community}}~\cite{Emsley:2016ws}.
    In this talk, Emsley discussed efforts to develop the community around the
    Drupal content management system, which is supported through the DrupalCon
    series of conferences, user groups, and an association.
    The talk described the use of social coding and software credit to develop
    the Drupal software community, and social documentation to provide
    documentation and tutorials. By analyzing identities of committers, they
    found that the larger community is actually a loosely coupled set of
    heterogeneous communities that use (and sustain themselves) through different
    social mechanisms.

    \item \textbf{Jack Dongarra, Sven Hammarling, Nicholas J.\ Higham, Samuel
    D.\ Relton\textsuperscript{\textasteriskcentered}, Pedro Valero-Lara,
    and Mawussi Zounon: \textit{Creating a Standardised Set of Batched
    BLAS Routines}}~\cite{Dongarra:2016ws}.
    This talk described Batched BLAS (BBLAS), a new standard for computing multiple
    BLAS (Basic Linear Algebra Subproblems) in parallel. Unlike existing parallel
    BLAS algorithms provided by, e.g., Intel MKL, cuBLAS, or MAGMA, BBLAS is
    optimized for smaller matrices, which apply to a variety of problems in areas
    such as machine learning, multifrontal solvers, image processing, fluid
    dynamics, and astrophysics. BBLAS offers nearly a factor of two improvement
    in computing speed compared to non-batched operations. To develop a sustainable
    BBLAS library, the developers have focused on proposing a standard API,
    holding workshops between academics and vendors to reach consensus, and
    performing research into API design decisions on performance.

    \item \textbf{Lukas Breitwieser\textsuperscript{\textasteriskcentered},
    Roman Bauer, Alberto Di Meglio, Leonard Johard,
    Marcus Kaiser, Marco Manca, Manuel Mazzara, Fons Rademakers,
    Max Talanov: \textit{The BioDynaMo Project}}~\cite{Bauer:2016ws}.
    Breitwieser described the BioDynaMo project, a large-scale platform for
    biological simulations that provides access to computational resources while
    hiding the complexity of distributed computing. It also promotes computational
    reproducibility through shared open-access data. The project began as a
    modernization initiative porting an existing simulation software, Cx3D, from
    Java to C++, but also incorporated many modern software engineering practices
    such as hosting and version control on GitHub, automated tests and continuous
    integration via Travis-CI, code coverage reports, and an automated build system.
    In addition, over 1000 students participated in development via the Intel
    Modern Code Developer Challenge competition---the winner achieved a
    performance speedup of 320 times. The project plans to continue improving
    the development process with (e.g.) static code checkers and Git commit hooks,
    continued architecture redesign, and verification.

    \item \textbf{Aseel Aldabjan, Robert Haines, and
    Caroline Jay\textsuperscript{\textasteriskcentered}: \textit{How should we
    measure the relationship between code quality and software
    sustainability?}}~\cite{Aldabjan:2016ws}.
    This talk explored the relationship between code quality and the sustainability
    of research software; however, the latter has no concrete definition, thus making
    it challenging to provide guidance on how to achieve sustainability. The authors
    proposed a ``software sustainment'' metric based on the time difference between
    the most recent and initial commits\slash contributions to the codebase.
    Using this metric, they analyzed 3113 Java projects hosted on GitHub, created
    in 2009, and found that 22\% and 35\% had sustainment metrics of 0 and less
    than seven, respectively. Interestingly, the authors also found high
    code quality (measured using a variety of static analytic metrics) across
    all these projects, with little correlation to sustainability.


\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Best practices for scientific software panel discussion} \label{sec:panel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\note{from Simon}

To investigate best practices for scientific software we put together a panel of five people with different perspectives on ``Best Practices for Scientific Software''.
Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software.
Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester are both Research Software Engineers with decades of experience of writing code for researchers.
Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business.
Finally, Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.

The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that does not span the Atlantic.
After translating this into the more open ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved in, not just using, but also writing code.
Hence, the following discussions would \todo{what goes here?}

The panel opened by noting that the world of scientific computing is changing rapidly, and the panel asked which practices have emerged recently that, if not handled correctly, might be the next generation's problem?
This quickly led to a discussion about the dangers of containerization.
These technologies are pushed by some---a minority it is hoped---as a panacea for sustainability.
The panel and attendees generally agreed that containerization is useful, but should not be seen as the solution to sustainability.
After all, these technologies are just as liable as any to decay.

A later question looked at the common practices from software engineering that the scientific community should be using.
The discussion that followed picked up on the key elements of sustainability, version control, documentation and ensuring that communities are active and engaged.
When asked whether there was resistance from the research community to adopting these practices, few negative experiences were reported.
It would appear that the problem lies in getting the message out about these technologies rather than trying sell researchers on the benefit of them.

The panel rounded up with a final question about "If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?

\todo{} <WILL ADD HERE ONCE I'VE TALKED TO PEOPLE>


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Working groups} \label{sec:WGs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

After most of the lightning talks and other presentations, WSSSPE used three areas of a large room to let
attendees use sticky flip charts and sticky notes on the walls to suggest
\begin{enumerate}
\item A vision on any aspect of the work.
\item A gap or challenge
\item A project idea
\end{enumerate}
(with one area of the room for each topic.)

Once we had the walls covered in ideas, we had people organically form groups around the flip charts on the wall\footnote{Photos of the flipcharts: \url{https://github.com/WSSSPE/media/tree/master/WSSSPE4-photos}}. For those folk who worked on Vision and Gaps, after an hour, we encouraged them to finish their Vision or Gap Google Doc, and to find and join or start working on a project with a group. This led to the 12 groups that are discussed in the rest of this section, specifically:
\begin{enumerate}
\item Verifying best practices \& metrics for sustainable research software
\item Software Sustainability Alliance
\item Scientific Software Prototyping Infrastructure (S2PI)
\item Standard metadata for software (CodeMeta)
\item White paper on developing sustainable software
\item Social science for scientific software
\item Software best practices for undergraduates
\item Meaningful metrics for sustainable software
\item Coordinating access to CI for research software
\item Software engineering processes tailored for research software
\item Open research index
\item Letters of evaluation for computational scientists
\end{enumerate}


The groups worked in Google docs, and eventually created Google slides to show their work back to the larger groups. The docs and slides were stored in a common Google Drive folder\footnote{\url{https://drive.google.com/drive/u/0/folders/0BzIpChHwv-q5X3lRaWhYMXJZaUE}}. In each group's discussion, a number of Specific, Measurable, Achievable, Relevant and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.  In many cases, this was done through Slack (see \S\ref{sec:slack}.)


%\todo{need to confirm who will write Access subsection (\S\ref{sec:access})}

\input{wg_best_practices_sustainability}
\input{wg_alliance}
\input{wg_prototyping}
\input{wg_codemeta}
\input{wg_best_practices_developing}
\input{wg_social}
\input{wg_best_practices_undergrads}
\input{wg_metrics}
\input{wg_access}
\input{wg_soft_eng}
\input{wg_open_research_index}
\input{wg_letters}

\section{Slack team and channels}\label{sec:slack}


As part of the discussion at WSSSPE4, we created a Slack team for WSSSPE~\cite{WSSSPESlack}.  To join it, go to \url{https://wssspe.signup.team/}.  A number of the working groups also created channels within the team; see the subsections in \S\ref{sec:WGs} for the names of these channels.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Attendee survey \label{sec:survey}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\note{from Lorraine}

All participants present on the last day of the meeting, September 14, 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
44 responses were recorded, with one respondent indicating that they took the survey twice.
The survey URL was also later distributed to attendees by email and advertised on twitter for attendees who were not there on the last day to have a chance to respond, and one additional response was recorded after the meeting.
Hence, the total number of unique individuals responding to the survey is likely 44.
The survey contained 9 multiple choice and open-text answer questions.
Text responses have been alphabetize to preserve anonymity.
The survey had a 100\% completion rate.
Appendix~\ref{sec:survey_details} contains the detailed survey results.

In general, the respondents were highly satisfied with the meeting and interested in continuing and participating in WSSSPE activities.
The mix of topics and types of interactions---talks, panels and discussion---was well balanced with several indicating a desire for time for questions and answers following talks as well for discussions in general.
Many respondents indicated that they will remain engaged in WSSSPE working group initiatives.
However, demand for a professional organization encompassing WSSSPE interests was weaker with over half interested in joining and a large percentage willing to consider the idea.
Respondents were grateful for the opportunity to network, explore new collaborations and for travel support to the conference.

The survey results indicate that future WSSSPE conferences should consider the balance of the attendees.
This year's meeting had many first time and early career participants who would have benefited from a review of basic concepts and terminology including what it means for software to be sustainable and the roles of research software engineers.
Additional emphasis and topics to explore would be the inclusion of more case studies, focus on the decision making process in developing and using software, deeper dives into selected topics, tutorials, software training within and outside of STEM, lowering the barriers to implementing best practices in software development, and progress towards executing WSSSPE's vision.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In WSSSPE4, we

\todo{discuss Twitter activity}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgments} \label{sec:acks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% NSF grant for WSSSPE4
This material is based upon work supported by the National Science Foundation (ACI-1648293),
by the Gordon and Betty Moore Foundation (GBMF\#5620),
and by the Alfred P.~Sloan Foundation (G-2016-7214).

\todo{feel free to add stuff here}

\newpage
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Organizing committee}  \label{sec:orgcom}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Do we want email addresses here?}

The following is the list of WSSSPE4 organizers.

{\scriptsize
\begin{longtable}{lll}
\input{orgcom}
\end{longtable}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Attendees}  \label{sec:attendees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The following is a list of attendees at WSSSPE4.

{\scriptsize
\begin{longtable}{lll}
\input{participants}
\end{longtable}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Travel award recipients}  \label{sec:awardees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Do we want email addresses here?}
The following table contains the list of WSSSPE4 travel award recipients, where
\textsuperscript{*} and \textsuperscript{\textdagger} indicate students and
early-career researchers, respectively.

{\scriptsize
\begin{longtable}{lll}
\input{awardees}
\end{longtable}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Program committee}  \label{sec:progcom}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The following table lists the WSSSPE4 program committee.

{\scriptsize
\begin{longtable}{lll}
\input{program_committee}
\end{longtable}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed survey results}  \label{sec:survey_details}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{survey_details}

%\input{wg_appendix_best_practices}

\newpage
\bibliographystyle{vancouver}

\bibliography{wssspe}
\end{document}
