%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Training Working Group Discussion}
\label{sec:appendix_training}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Nick Jones\footnote{email:
\href{mailto:nick.jones@nesi.org.nz}{nick.jones@nesi.org.nz}} will serve as the
point of contact for this working group, and be responsible for ensuring timely
progress of the planned actions.

\subsection{Group Members}
\begin{itemize}
\item Nick Jones -- New Zealand eScience Infrastructure
\item Iain Larmour -- Engineering \& Physical Sciences Research Council, UK
\item Erin Robinson -- Foundation for Earth Science
\end{itemize}

\subsection{Summary of Discussion}


While little training focuses specifically on this outcome \katznote{sustainable
software?}, a variety of training activities increase researcher awareness of
and engagement with software professionals and software engineering practices.
Research Software Engineers are being recognized as critical contributors to
high quality research; the pathway to acquire and master the relevant skills
is not yet clear; equally those skills required by researchers in general are
also not commonly understood nor routinely developed.

The group's discussion explored a rapidly growing array of training that is seen
to contribute to sustainable software. The offerings are diverse, including:
self-paced online modules focused around specific tools; single and multiple day
training workshops that raise awareness of a tool chain to support collaborative
and shared software development within a research workflow; block courses
specializing on particular methods, technologies, and applications; academic
programs at undergraduate and masters levels; doctoral training programs that in
part contain requisite skills training activities.

While some of this training focuses on applying software engineering practices
within the context of research, meeting the values and goals of research are
less often incorporated as explicit learning outcomes. With software (and
similarly, data) often the only tangible artifact of a research method or
protocol, the dependency between software applications and the quality of
research adds complexity to the learner's journey. In recognition of the longer
term investment required by researchers to integrate such skills into their
research practices, many activities are focusing on emotionally engaging
researchers and cohorts, to build a sense of shared purpose beyond the obvious
goal of technical skill acquisition.

In reviewing current training activities, the group identified a variety of
perspectives seen as useful in positioning activities in ways to better
communicate why and when best to apply each activity. Training can be
categorized on a variety of spectra, with content and delivery ranging across them, for example:
programming to research; basic to advanced; technical to emotional; informal to
formal; and self-paced to participative. A few attempts have been made to situate a
cross section of training activities within such dimensions, creating easier
means of communicating the value of any specific opportunity and the pathways
across opportunities over time.

Evaluation of training delivery and outcomes is seen as a weakness common to
most non-academic training activities. Opportunities for measuring success in
delivering training start simply with collecting a Net Promoter Score, which
lets those delivering training know whether attendees are likely to recommend
the training to others. In looking at the longer term outcomes for the learner,
frameworks such as Bloom's taxonomy and Kirkpatrick's evaluation model offer possible
approaches.

In this latter case of formal evaluation, ownership of evaluation as a component
of career development for any researcher appears mostly absent. While academic
research institutions have professional development centers to support research
staff, the skills taught which might impact on sustainable software are limited
at best, and lack a clear and coherent development pathway.

Coordination of these training projects will depend on buy-in from a broad range
of training program and activity leaders, suggesting a key opportunity lies in
identifying and bringing together these people on a regular basis.

\subsection{Description of Opportunity, Challenges, and Obstacles}

Software skills are needed by an increasing array of researchers and fields. The
training arc is not well-defined, with a sometimes baffling array of training
opportunities responding to various facets of skill deficit and need. Given this
current complexity, coordination across training projects would create common
frames of reference, communicating and integrating activities to better serve
the needs of researchers.

Building this community could lift the maturity of training projects and
capabilities, enabling more advanced approaches to address key gaps in
evaluation, career development, and a lift in the standard of research
practices.

In aiming at these opportunities, it will be necessary to find the means to
support those involved in leading training activities to allocate time to
coordination activities, which will often sit beyond their current scope of
responsibility.

These activities are also distributed globally, with no single country or region
offering a comprehensive set of capabilities and initiatives. Any coordination
activity will therefore need to raise the profile of the opportunity gap with
relevant research funders and policy makers.

\subsection{Key Next Steps}

The goal of the following next steps is to quickly test whether there is
interest in establishing a community committed to increasing the degree of
coordination across training projects.

\begin{enumerate}

\item Hold a virtual meeting by December 2015, to bring together a broader group
of interest in this topic, with specific goals to:

	\begin{enumerate}
	    
	\item Identify programs with existing activities aimed at integrating across
	training projects.
	        
	\item Identify training projects with an interest in participating in
	coordination efforts.
	        
	\item Identify funding opportunities to bring together training program and
	project leaders to identify shared goals for future coordination of activities.
	        
	\item Agree on a communications plan to qualify whether programs, projects, and
	funders are interested in engaging and committing to ongoing activities.
	        
	\end{enumerate}
    
\item Review progress within 3 months, to establish next steps, if any.

\end{enumerate}

\subsection{Plan for Future Organization}

Continue to track progress by posting comments to WSSSPE3 issue.

\subsection{What Else is Needed?}

If the group moves from early-stage formation into working towards shared goals,
expertise will likely be required in pedagogy and training evaluation.

\subsection{Key Milestones and Responsible Parties}
\begin{enumerate}

\item During October, Nick Jones and Erin Robinson to draft WSSSPE3 report back.

\item Before December 2015, Nick Jones and Erin Robinson to call a meeting of
the broader group, to review key next steps.
    
\item First quarter 2016 -- if willing parties are identified, draft workshop proposal
and identify a relevant forum, including future WSSSPE events.
    
\end{enumerate}

\subsection{Description of Funding Needed}

Workshop/RCN travel funding to bring together key program, project, and funder
representatives from across North America, EU, UK, Australasia. In addition,
funding to support work on better defining the landscape of training activities,
the useful perspectives in communicating the value of the varied training
projects, and the possible pathways through training activities over time.
