\subsection{Topic} \todo{Training Working Group Discussion}

\subsubsection{Why it is important}

This group explored a rapidly growing array of training which is seen to contribute to sustainable software. Offerings are diverse, providing training more or less directly relevant to sustinable software. While research institutions  support professional development for research staff, the skills taught which might impact on sustainable software are limited at best, often lacking a clear and coherent development pathway. This growing array of training opportunities could usefully be coordinated, by bringing together those involved in leading relevant initiatives, on a regular basis.

\subsubsection{Fit with related activities}
Two existing venues for discussion of related activities are identified:

\begin{itemize}

\item
\item
WSSSPE Workshop on Sustainable Software for Science: Practice and Experiences \footnote{\url{http://wssspe.researchcomputing.org.uk/}}

\item
SEHPCCSE International Workshop on Software Engineering for High Performance Computing in Computational Science and Engineering \footnote{\url{http://wssspe.researchcomputing.org.uk/}}

\end{itemize}

\subsubsection{Discussion}

Next steps have been identified to quickly test whether there is interest in establishing a community committed to increasing the degree of coordination across training projects.

See Appendix~\ref{sec:appendix_training} for more details about the discussion.

\subsubsection{Plans}

The main plans for the group are to convene discussion to explore bringing together a regular meeting of those involved in leading relavent training projects.

\subsubsection{Landing Page}

The Training working group requests an email be sent to Nick Jones \href{mailto:nick.jones@nesi.org.nz}{(nick.jones@nesi.org.nz)} to find out more about the group's efforts and how to participate.
